---
title: "R Code Basics"
subtitle: "BDiB Session 11-February-2021"
output: html_notebook
---

<style>
h1 {background: brown;color: white;padding-left: 7px;}
h2 {background: grey;color: white;padding-left: 7px;}
h3 {color: brown;}
</style>


## Scripts  
***ALWAYS SAVE YOUR CODE IN SCRIPTS !***  

* To document your work (think of them as a data analyst's lab notebook)
* To quickly repeat an analysis in the future
* To re-use your code for new analyses
* To generate reports (rmarkdown scripts)

There are two types of R scripts:  

* Standard R scripts (see `basics.R`)  
* rmarkdown (`.Rmd`) scripts  
* Please see rmarkdown.Rmd or rmarkdown.nb.html for details on rmarkdown scripts  

## Data structures (1): scalars and vectors  

* Vectors are simple lists with each item being of the same type.
* Scalars are really just vectors with a single item.
* Some useful vectors are 'built' into R:
* (place the cursor anywhere in a line of code and press Ctrl-Enter to execute
* the code / Cmd-Enter on a Mac)
* Results / output will appear in the 'Console' window below the script window.

```{r}
letters
```

```{r}
LETTERS
```

```{r}
month.abb
```

```{r}
month.name
```


Vectors can be generated using the c() function:
```{r}
c(14, 6, 2016, 12)
```

## Functions  
***Functions are used to manipulate data*** 

### Standard arithmetic  
R includes all the standard arithmetic function symbols: + - * / ^  
```{r}
3 + 4
```

```{r}
300 / 100
```

```{r}
10 - 9
```

```{r}
2 ^ 4
```

Modulo:  
```{r}
12 %% 5
```

### Names functions  
Named R functions are very similar to worksheet functions in Excel:  

1. Each function has a different name  
2. Function names are always followed by brackets  
3. Information typed between the brackets are called "arguments"  
4. Two or more arguments are separated by commas  

To see a list of all (>1200) built-in functions that come with R (aka "base R") type:  
```{r}
help(package="base")
```

So `help()` is a first example of a function. ' package="base" ' is the argument here.

### Some examples of frequently used functions  
(note how in some the examples we use vectors as arguments)  

```{r}
c(3, 4, 10)
```

```{r}
sum(c(3, 4, 10))
```

```{r}
sqrt(121)
```

```{r}
mean(c(3, 4, 10))
median(c(3, 4, 10))
max(c(3, 4, 10))
min(c(3, 4, 10))
```

