---
title: "R Code Basics"
subtitle: "BDiB Session 11-February-2021"
output: html_notebook
---

<style>
h1 {background: brown;color: white;padding-left: 7px;}
h2 {background: grey;color: white;padding-left: 7px;}
h3 {color: brown;}
</style>


## Scripts  
***ALWAYS SAVE YOUR CODE IN SCRIPTS !***  

* To document your work (think of them as a data analyst's lab notebook)
* To quickly repeat an analysis in the future
* To re-use your code for new analyses
* To generate reports (rmarkdown scripts)

There are two types of R scripts:  

* Standard R scripts (see `basics.R`)  
* rmarkdown (`.Rmd`) scripts  
* Please see rmarkdown.Rmd or rmarkdown.nb.html for details on rmarkdown scripts  

## Data structures (1): scalars and vectors  

* Vectors are simple lists with each item being of the same type.
* Scalars are really just vectors with a single item.
* Some useful vectors are 'built' into R:
* (place the cursor anywhere in a line of code and press Ctrl-Enter to execute
* the code / Cmd-Enter on a Mac)
* Results / output will appear in the 'Console' window below the script window.

```{r}
letters
```

```{r}
LETTERS
```

```{r}
month.abb
```

```{r}
month.name
```


Vectors can be generated using the c() function:
```{r}
c(14, 6, 2016, 12)
```

## Functions  
***Functions are used to manipulate data*** 

### Standard arithmetic  
R includes all the standard arithmetic function symbols: + - * / ^  
```{r}
3 + 4
```

```{r}
300 / 100
```

```{r}
10 - 9
```

```{r}
2 ^ 4
```

Modulo:  
```{r}
12 %% 5
```

### Names functions  
Named R functions are very similar to worksheet functions in Excel:  

1. Each function has a different name  
2. Function names are always followed by brackets  
3. Information typed between the brackets are called "arguments"  
4. Two or more arguments are separated by commas  

To see a list of all (>1200) built-in functions that come with R (aka "base R") type:  
```{r}
help(package="base")
```

So `help()` is a first example of a function. ' package="base" ' is the argument here.

### Some examples of frequently used functions  
(note how in some the examples we use vectors as arguments)  

```{r}
c(3, 4, 10)
```

```{r}
sum(c(3, 4, 10))
```

```{r}
sqrt(121)
```

```{r}
mean(c(3, 4, 10))
median(c(3, 4, 10))
max(c(3, 4, 10))
min(c(3, 4, 10))
```
## Variables  
Usually data are stored in named VARIABLES.  

* Variables are defined with the '<-' assignment operator  
* Variable names can be very long  
* Variables CANNOT contain spaces  
* Variables CANNOT contain dashes/hyphens    
* Variables CANNOT start with a number (but numbers can appear elsewhere)  

# Two scalars:
```{r}
my.first.variable <- "Hello world!"
my.second.variable <- 2019
```

***Note that variables are listed in the 'Environment' tab in the top right RStudio panel.***  

Use the 'print()' or 'cat()' function to display the contents of a variable.  
Output from `print()` is preceded by row numbers in square brackets; with `cat()` the row numbers are not printed:
```{r}
print(my.first.variable)
cat(my.first.variable)
```

```{r}
print(my.second.variable)
cat(my.second.variable)
```

Vectors can be generated using the `c()` function:
```{r}
v1 <- c(14, 6, 2016)
```

Typing just the variable is equivalent to `print()`:
```{r}
v1
```

The `View()` function opens a new tab - mostly used to inspect large tables.  
The tab will include several useful ways of inspecting and filtering a table.
```{r}
View(mtcars)
```

## Playing with vectors  
**Note**: The objective section is just to illustrate the usefulness of functions and programming languages in general. You do ***not*** need to remember any of the specifics!  

### Playing with text vectors  
```{r}
v2 <- c("Hello", "world", "!")
v2
```

Vector items can be 'glued' together into a single piece of text using the `paste()` function:
```{r}
paste(v2, collapse=" ")
```

Modifying capitalization:  
```{r}
toupper(v2)
tolower(v2)
```

### Playing with number vectors  
It's easy to generate vectors with simple series of numbers:

```{r}
v3 <- 1:12
v3
```

The seq command can generate more complicated number vectors:
```{r}
seq(from=1, to=22, by=3)
```

```{r}
vv <- seq(from=1, to=10, length.out=19)
vv
```

The `length()` function returns the number of items in a vector:  
```{r}
length(vv)
```

### Plotting vectors  
Plotting vectors is very straightforward:  
```{r}
plot(1:50)
```
```{r}
t <- (1:100) / 10
plot(sqrt(t) * cos(t), sqrt(t) * sin(t), type="l", axes=FALSE, col="blue")
```


### Vector items can have names  
The `names()` function is used both to define and to return the names of a data object.
```{r}
v3
month.abb
```

```{r}
names(v3) <- toupper(month.abb)
v3
```

```{r}
names(v3)
```

In R it is very easy to manipulate all items in a vector with simple code:  
```{r}
v3 * 10
```

## Subsetting (1): Subsetting vectors  

### Subsetting vectors by index  

You can select items from within a vector using square brackets + indices ...  

```{r}
# LETTERS is a built-in variable that stored the uppercase letters of the alphabet:
LETTERS
```


Subsetting with a single index number...  
```{r}
LETTERS[11]
```

Subsetting with a vector of index numbers ...  
```{r}
# The c() function generates a vector:
c(8,5,12,12,15)
```

```{r}
h <- LETTERS[c(8,5,12,12,15)]
h
```

Remember how to 'glue' vector items together?
```{r}
paste(h, collapse="")
```

### Subsetting vectors using names  
You can also refer to vector items by name, if defined ...
```{r}
v3
```

```{r}
# v3[11]
v3["NOV"]
# v3["NOV"] * 10000
```

Using a vector of names:
```{r}
# c("JAN", "MAR")
# v3[c(1,3)]
v3[c("JAN", "MAR")]
```

### Subsetting vectors using Boolean (TRUE/FALSE) values  
```{r}
v4 <- 1:3
v4
```

```{r}
c(TRUE, FALSE, TRUE)
v4[c(TRUE, FALSE, TRUE)]
```

It can be easier to save a vector in a variable first, then use the variable for subsetting:
```{r}
v4
v4 > 1
x <- v4 > 1
x
v4[x]
```
## Data types in R  

As we have already seen the most important data TYPES in R are:  
1. numeric  
2. text (aka 'character' or 'strings')  
3. Boolean (aka 'logical')  

*(numeric data are split further into sub-categories but we don't to worry
about that here)*  

### Numerical data  
```{r}
v3
is.numeric(v3)
class(3.5)
# typeof(v3) # integer = whole numbers
# typeof(3.5) # "double"
```

### Text data  
Text format in R is called 'character':  
```{r}
v2
class(v2)
# is.character(v2)
# is.numeric(v2)
# typeof(v2)
```

### Boolean/logical data  
Boolean/logical values are often generated through 'equal' or 'greater/lesser than' (<==>) operations:  

```{r}
10 > 2
2 > 10
3 == 3.0
4 == 5
```
The following generates a vector of boolean values:
```{r}
1:10         # generates a vector of numbers from 1 to 10
1:10 >= 5    # generates a vector of boolean values
```

Store in a vector:  
```{r}
v5 <- 1:10 >= 5
v5
```

To check and illustrate let's add names for the vector items:
```{r}
names(v5) <- paste("No.", 1:10, sep="")
v5
```

Boolean values in R are called "logical":  
```{r}
class(v5)
is.logical(v5)
# is.logical(letters)
# typeof(letters)
# is.numeric(1:10)
# is.numeric(1.2)
```

## Data structures (2): Matrices  

Matrices are 2-dimensional arrays/tables where each item (think of Excel cells) has to be of the same data type:

```{r}
# generates a vector
1:24
```

Convert vector into a matrix with 6 columns  
```{r}
mx1 <- matrix(1:24, ncol=6)
mx1
```
```{r}
# Fill by column:
# mx2 <- matrix(LETTERS[1:24], nrow=6, byrow=FALSE)
# mx2
```

```{r}
# Fill by row:
# mx3 <- matrix(LETTERS[1:24], nrow=6, byrow=TRUE)
# mx3
```

## Calculations with vectors + matrices  

As we have already seen, one of the great things about vectors and matrices in R is that they can be modified with simple code:

```{r}
# v3
v3 * 100
```
```{r}
# mx1
mx1 + 1000
```
```{r}
# Transpose a matrix with 't()' function:
# mx1
# t(mx1)
```

## Data structures (3): Data frames  

Data frames are the most common container used to store data in R.  
Each column can be a different data type:  

```{r}
# 'iris' is another variable 'built into' R that stores a data frame:
# class(iris)
# class(mx1)
```

GEenrating a data frame 'by hand':
```{r}
df1 <- data.frame(Name=c("Jane", "Jack", "Jaim"),
                  Age=c(5, 12, 30),
                  Female=c(TRUE, FALSE, FALSE))
df1
```

To show what data type is found in each column of the data frame use this code:
```{r}
sapply(df1, class)
```

Very large data frames are best viewed in a separate tab:  
```{r}
View(df1)
```



